---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# ahpsurvey

Date: 8 September 2018    
Version: 0.1.0

## Overview

The `ahpsurvey` package provides a consistent methodology for researchers to reformat data and run the analytic hierarchy process (AHP) on data that are formatted with the survey data entry mode. It is optimised for performing the AHP with many decision-makers, and provides tools and options for researchers to aggregate individual preferences and concurrently test multiple aggregation options. It also allows researchers to quantify, visualise and correct for inconsistent pairwise comparisons.

## Installation

You can install `ahpsurvey` from Github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("frankiecho/ahpsurvey")
```
A CRAN submission is under way and a direct CRAN download will be available in the near future.

## Usage

Here, we have the simulated survey data of pairwise comparisons of 200 decision-makers who responded to a survey about the attributes they think is important when choosing a city to live in.

```{r example}
library(ahpsurvey)
library(magrittr)

data(city200)
city200 %>% head()
```

`ahpsurvey` allows us to convert this `data.frame` into a usable list of pairwise comparison matrices for our further action:

```{r}
## Define the attributes used
atts <- c("cult", "fam", "house", "jobs", "trans")

city200 %>%
  ahp.mat(atts = atts, negconvert = TRUE) %>%
  head(2)
```

And can calculate the aggregated individual preferences of the 200 decision-makers using that list:

```{r}
city200 %>%
  ahp.mat(atts = atts, negconvert = TRUE) %>%
  ahp.aggpref(atts = atts)
```

Further arguments allow you to specify the aggregation method, impute missing values and identify and correct inconsistent responses.

## Functions

An overview of the functions in this package are as follows:

* `ahp.mat`: Generate AHP pairwise matrices from survey data
* `ahp.indpref`: Priority weights of individual decision-makers
* `ahp.aggpref`: Aggregate individual priorities (AIP)
* `ahp.aggjudge`: Aggregate individual judgements (AIJ)
* `ahp.cr`: Saaty's Consistency Ratio
* `ahp.error`: The product between the pairwise comparison value and pj/pi
* `ahp.pwerror`: Finds the pairwise comparisons with the maximum amount of inconsistency
* `ahp.missing`: Impute missing pairwise comparsions
* `ahp.harker`: Replace inconsistent pairwise comparisons

## Vignettes

For a detailed example of how the above function works, look no further than the vignettes, which are stored in `/vignettes/my-vignette.md/`. There, you can find a detailed step-by-step instruction of how to use the function using a simulated survey dataset and visualise the output using `ggplot2`.

## Future development

I have plans to add the following features in the future, perhaps after I finish writing up my masters thesis :-(

* Multiple level of attributes: right now, you can always multiply the weights manually, but I'm looking to develop this feature in a convenient function
* Comparing alternatives: or a way to export the matrices to be used in other packages which does this
* Sensitivity analysis
* More ways to impute missing data

Please let me know if there are any features which could be useful to you in a feature request or contribution.

## Author

* **Frankie Cho** - *Author & Maintainer* - [frankiecho](https://github.com/frankiecho)

## License

This project is licensed under the MIT License.